---
title: "barbac · Chen barcode time series"
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    toc-depth: 2
    toc-location: left
    theme: flatly
    page-layout: full
    code-fold: true
    smooth-scroll: false
    html-math-method: plain
engine: knitr
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
here <- getwd()
root <- normalizePath(file.path(here, "..", "..", ".."))
devtools::load_all(root, quiet = TRUE, export_all = FALSE)
source(file.path(here, "report_helpers.R"))
cache_file <- Sys.getenv("BARBAC_REPORT_DATA", file.path(here, "..", "generated", "r_report", "report_data.rds"))
report <- readRDS(cache_file)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 5.5, dpi = 130, dev = "ragg_png")
fmt <- function(x) format(round(x), big.mark = ",", scientific = FALSE, trim = TRUE)
lv <- report$summary[report$summary$method == "barbac_lv", ]
ham <- report$summary[report$summary$method == "barbac_hamming", ]
total <- sum(report$extraction$total_pairs)
molecules <- sum(report$extraction$barbac_input_molecules)
img <- function(path, alt) htmltools::tags$img(src = base64enc::dataURI(file = path, mime = "image/png"),
  alt = alt, style = "width:100%;height:auto;")
```

<style>
html { scroll-behavior:auto !important; }
body { color:#263544; font-size:16px; line-height:1.65; }
.main-container { max-width:1450px; }
h1.title { color:#2C4A63; margin-bottom:1.2em; }
h2 { color:#2C4A63; border-bottom:1px solid #dae1e7; padding-bottom:.4em; margin-top:2em; }
details { border:1px solid #d8e0e7; border-radius:5px; margin:12px 0; padding:12px 16px; }
summary { cursor:pointer; font-weight:600; color:#2C4A63; }
.metric-grid { display:flex; flex-wrap:wrap; gap:14px; margin:24px 0; }
.metric { flex:1; min-width:185px; padding:18px; background:#f3f6f8; border-left:4px solid #2C4A63; }
.metric strong { display:block; font-size:1.8em; }
.metric span { display:block; color:#52616c; font-size:.9em; }
.dataTables_wrapper { font-size:13px; margin:20px 0 30px; }
.nav-tabs { margin:20px 0; }
.nav-tabs > li > a { padding:10px 16px; }
.note { border-left:3px solid #B07D22; padding:10px 18px; background:#fcfaf4; margin:20px 0; }
.source { color:#596b76; font-size:.88em; }
.girafe { margin:12px 0 24px; }
@media(max-width:768px) { body {font-size:15px;} .main-container {padding:0 14px;} }
@media print { details {break-inside:avoid;} .tab-content>.tab-pane {display:block!important;} }
</style>

## Strong agreement with the published time series

**All eight samples are complete.** The shared workflow retains
`r fmt(molecules)` UMI-deduplicated molecules from `r fmt(total)` read pairs.
Barbac LV + Poisson reaches median Spearman agreement **`r sprintf("%.6f", lv$spearman_median)`**
with the published barcode counts. Both barcode components take **`r sprintf("%.2f", lv$clustering_seconds_median)` seconds**
to cluster at the median of three serial repeats. Hamming is faster by median;
Starcode agrees slightly more closely with the publication.

**This report is produced in R using barbac's own statistics and plotting functions.**
Use the sample panels, interactive composition tabs, searchable count tables,
and CSV download buttons to inspect the evidence. The HTML contains its figures
and interactive libraries and can be opened locally without an R server.

```{r headline}
htmltools::div(class = "metric-grid",
  htmltools::div(class = "metric", htmltools::strong(sprintf("%.2f%%", 100*molecules/total)),
    htmltools::span("Input read pairs retained as molecules")),
  htmltools::div(class = "metric", htmltools::strong(sprintf("%.2f%%", lv$published_mass_percent_median)),
    htmltools::span("Median LV molecule coverage on exact published pair IDs")),
  htmltools::div(class = "metric", htmltools::strong("4 × 2"),
    htmltools::span("Measured generations × biological replicates")))
```

The assay is **hBFA1 / YPD**, sampled at generations **8, 16, 24 and 40** in
replicates **R1 and R2**. A lineage identifier is the ordered barcode pair
**BC2_BC1**. Each component is normally 26 bases; observed lengths 24–28 are
retained. Read pairs, merged alignments, distinct sequences, clusters, and
UMI-deduplicated molecules are different units and are labeled below.

<div class="note">The publication is a differently processed experimental reference,
not a set of true barcode labels. Retention, mapping percentages, and published-ID
coverage are not accuracy percentages. We cannot derive biological FP/FN from this comparison.</div>

## Mapping retains most merged reads

The mapping overview comes directly from `summarise_bam_stats()` and
`plot_bam_stats()`. These counts concern **PEAR-merged BAM alignments**; the mapping
percentage uses mapped plus unmapped merged alignments as its denominator.
It excludes original pairs that PEAR did not merge. The table exposes the
original input-pair count for comparison.

```{r mapping-plot}
mapping_data <- report$mapping
mapping_data$sample <- factor(mapping_data$sample, levels = rev(report$manifest$sample))
mapping_plot <- barbac::plot_bam_stats(mapping_data) + barbac::theme_barbac(base_size = 12, family = "Arial") +
  ggplot2::labs(y = "Merged BAM alignments")
report_plotly(mapping_plot, tooltip = c("x", "y", "fill"))
```

```{r mapping-table}
x <- report$mapping
names(x) <- c("Sample", "Mapped", "Unmapped", "Input pairs", "Merged alignments", "Mapped / merged (%)")
report_table(x, "Counts recalculated from all eight completed BAMs.")
```

## Extraction retains observed barcode lengths

The following are the **actual panels returned by `barbac_xtr.stats()`** for both
components in each sample. Expand a sample to inspect barcode length, abundance,
nucleotide-composition entropy, and the length-bin summary. The requested range
is 24–28 bases. A regression test protects the corrected range calculation
against changes in input row order.

The histograms count **distinct extracted sequences**, each contributing once.
In these package panels, **“Reads” means retained UMI-deduplicated molecules**.
Sequence entropy measures nucleotide composition within a barcode, not population
diversity or clustering accuracy. These panels describe the retained input;
the filtering table below describes molecules removed earlier.

```{r extraction-panels}
htmltools::tagList(lapply(seq_len(nrow(report$manifest)), function(i) {
  sample <- report$manifest$sample[i]
  panels <- Filter(function(p) p$sample == sample, report$panels)
  htmltools::tags$details(open = if (i == 1) "open" else NULL,
    htmltools::tags$summary(paste(report$manifest$replicate[i], "· generation", report$manifest$generation[i], "·", sample)),
    htmltools::tagList(lapply(panels, function(p) htmltools::tagList(
      htmltools::tags$h4(p$component), img(p$path, paste(sample, p$component, "barbac extraction diagnostics"))))))
}))
```

```{r extraction-losses}
x <- report$extraction[c("sample", "total_pairs", "quality_failed", "no_complete_BAM_barcode_pair", "umi_duplicates", "barbac_input_molecules")]
x$`Retained (%)` <- 100*x$barbac_input_molecules/x$total_pairs
names(x)[1:6] <- c("Sample", "Input pairs", "Quality failed", "No complete BAM pair", "Repeated UMI", "Retained molecules")
report_table(x, "Sequential filtering outcomes; columns reconcile to the original input pairs.", digits = 2)
```

## Cluster statistics describe abundance, not truth

`cluster_stats()` summarizes the saved LV and Hamming cluster outputs.
**BC2 and BC1 are clustered separately** across all eight samples, then paired
identities are reconstructed. The component table therefore has the same pooled
molecule total in every row: it must not be added across components or modes.
Fewer clusters can reflect correction or over-merging; these statistics alone
cannot distinguish the two.

```{r component-clusters}
x <- report$cluster_components
x$method <- unname(report_labels[x$method])
names(x)[names(x) == "total_reads"] <- "total_molecules"
report_table(x, "Pooled component summaries from cluster_stats(). Fractions range from 0 to 1.", digits = 4)
```

For paired lineages, `cluster_stats()` is applied separately to each sample's
**positive counts**. The complete time-series tables also contain explicit zero
observations; these zeros are preserved for trajectories and excluded when
counting observed clusters. “Singleton” here means a lineage supported by one molecule.

```{r paired-clusters}
x <- report$cluster_pairs
x$method <- unname(report_labels[x$method])
names(x)[names(x) == "total_reads"] <- "total_molecules"
report_table(x, "Observed paired lineages per sample; fraction columns range from 0 to 1.", digits = 4, filter = "top")
```

## Major barcode pairs change across generations

These interactive panels are generated by **`barbac_ts_area(interactive = "ggiraph")`**.
Hover to identify a lineage and inspect its frequency; the toolbar supports zoom
and image export. The same four largest pooled LV pairs are shown in every panel,
with every other assigned pair retained in **Other**. This preserves the denominator
and gives the same identifiers the same colors across modes and replicates.

Frequencies use **all assigned molecules**, including unpublished pairs. Actual zeros
remain zero: `min_total_count = 0`, `fill_missing = "zero"`, and no time shift.
Areas connect only four measured generations; intermediate frequencies are visual
interpolation, not additional observations or fitted growth. The lookup table after
these panels gives the complete sequences behind the short labels.

::: {.panel-tabset}
### LV · R1

```{r lv-r1}
report$compositions[["barbac_lv R1"]]$widget
```

### LV · R2

```{r lv-r2}
report$compositions[["barbac_lv R2"]]$widget
```

### Hamming · R1

```{r ham-r1}
report$compositions[["barbac_hamming R1"]]$widget
```

### Hamming · R2

```{r ham-r2}
report$compositions[["barbac_hamming R2"]]$widget
```

:::
## Every paired count remains available

Search for a full barcode pair or either sequence component, sort by abundance,
and download the filtered table as CSV. These tables contain the **complete barcode
universe for each barbac mode**, including explicit zero counts and published IDs
that were not observed. Each generation column contains UMI molecule counts;
the pooled column sums the eight samples. A published-ID match identifies exact
sequence membership, not validated biological truth.

::: {.panel-tabset}
### Plot label lookup

```{r label-lookup}
report_table(report$lookup, "Full pair identifiers used in the composition and publication figures.", digits = 0)
```

### LV counts

```{r lv-explorer}
report_table(report$explorer$barbac_lv, "Complete LV barcode-pair counts across all eight samples.", digits = 0, page_length = 8)
```

### Hamming counts

```{r ham-explorer}
report_table(report$explorer$barbac_hamming, "Complete Hamming barcode-pair counts across all eight samples.", digits = 0, page_length = 8)
```

:::
## Publication agreement is high, with unmatched mass

**LV closely reproduces the ranking of published barcode counts.** The correlation
uses all 2,314 published pair IDs in each sample, inserting zero for absent calls.
All 48 correlations, coverage percentages, and frequency differences were
recalculated in R and checked against the completed comparison.

The figure shows each sample's correlation. Its focused correlation range makes
differences near one visible; it is not a percentage axis. The exact median values
and retained molecule coverage appear in the table. Starcode is slightly closer
to the reference on this assay; that does not establish superior biological accuracy.

```{r agreement-plot}
x <- report$agreement
x$Method <- factor(unname(report_labels[x$method]), levels = unname(report_labels))
p <- ggplot2::ggplot(x, ggplot2::aes(Method, spearman, text = sample)) +
  ggplot2::geom_point(position = ggplot2::position_jitter(width = .09, height = 0, seed = 42),
    color = report_blue, alpha = .7, size = 2) +
  ggplot2::coord_flip() + ggplot2::labs(x = NULL, y = "Spearman correlation (focused range)",
    title = "Count-rank agreement across eight samples") + barbac::theme_barbac(base_size = 12, family = "Arial")
report_plotly(p, tooltip = c("text", "x", "y"))
```

```{r method-table}
x <- report$summary[c("method", "spearman_median", "published_mass_percent_median", "clustering_seconds_median", "unassigned_molecules")]
x$method <- unname(report_labels[x$method])
names(x) <- c("Method", "Median Spearman", "Median published-ID molecule coverage (%)", "Median clustering (s)", "Unassigned molecules")
report_table(x, "Agreement/coverage: medians across eight samples. Time: median of three repeats.", digits = 6)
```

The **`r sprintf("%.2f%%", lv$published_mass_percent_median)`** LV coverage value is
the median across samples. Weighting all samples by their molecule counts instead,
**`r sprintf("%.2f%%", report$unmatched$pooled_unmatched_percent)`** of LV molecules
fall outside the published pair set. Two abundant absent pairs account for
**`r sprintf("%.2f%%", report$unmatched$largest_two_percent_of_unmatched)`** of this
unmatched mass. Both have BC2 at least seven edits from every published BC2,
outside the distance-three correction radius. Their biological validity and the
reason for their absence from the author reference remain unresolved.

## Clustering time is separate from preprocessing

Hamming has the lowest median clustering time in this run. LV has a similar
median time to Starcode. Bars show **medians of three serial repeats**, and
whiskers show their **minimum–maximum range**. These are observed ranges, not
confidence intervals. All three repeats produced identical memberships.

```{r timing-plot}
x <- report$summary
x$Method <- factor(unname(report_labels[x$method]), levels = unname(report_labels[x$method[order(-x$clustering_seconds_median)]]))
p <- ggplot2::ggplot(x, ggplot2::aes(Method, clustering_seconds_median)) +
  ggplot2::geom_col(fill = report_blue, width = .65) +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = clustering_seconds_min, ymax = clustering_seconds_max), width = .2) +
  ggplot2::coord_flip() + ggplot2::labs(x = NULL, y = "Combined BC2 + BC1 workflow (seconds)",
    title = "Clustering workflow time", subtitle = "Startup, clustering and required exports; shared preprocessing excluded") +
  barbac::theme_barbac(base_size = 12, family = "Arial")
report_plotly(p, tooltip = c("x", "y", "ymin", "ymax"))
```

Timing includes process startup, clustering, required format conversion and
centroid/member exports for both pooled components. It excludes downloading,
FastQC, PEAR, mapping, barcode extraction, trajectory reconstruction, comparison,
and creation of this report. The benchmark ran serially on macOS ARM64 with
16 GiB RAM; method thread settings were one. These times are not predictions
for a different cluster or sequencing depth.

## Methods and validation

The workflow is FastQC → PEAR → minimap2 → indexed BAM → `barbac_xtr()` flank
extraction → original-pair quality filtering → original-order UMI deduplication.
Both barcode components retain their observed sequence lengths and their pair
identity; BC1 is converted to the author's orientation. PEAR consensus can change
the barcode relative to the original mate, which the extraction audit records.

All methods receive the same molecule counts at **distance 3**. Barbac uses
support ordering, merge ratio 20, configured error rate 0.005, and the optional
Poisson indel model for LV only. Shepherd uses supplied error rate 0.005 after
automatic estimation failed on BC1. That choice preceded publication comparison.
The error-rate setting is an assumption, not an estimated error rate for these reads.

The report recomputes mapping statistics, checks extracted component totals,
calls `cluster_stats()`, and recalculates correlations and timing summaries in R.
The saved clustering assignments and original comparison files are unchanged.
Original FASTQ sizes/MD5s, FastQC counts, frequency denominators and repeated
memberships passed the preceding complete-analysis validation.

```{r checks}
data <- data.frame(Check = gsub("_", " ", names(report$checks)),
  Result = ifelse(unlist(report$checks), "Passed", "Failed"))
knitr::kable(data, row.names = FALSE)
```

<details><summary>R functions used in this report</summary>

```{r function-examples, echo=TRUE, eval=FALSE}
mapping <- summarise_bam_stats(bam_dir)
plot_bam_stats(mapping)
barbac_xtr.stats(extracted_component, barcode_length = c(24, 28))
cluster_stats(component_centroids)
cluster_stats(data.frame(sum_counts = paired_counts[paired_counts > 0]))
barbac_ts_area(composition_counts, min_total_count = 0,
               fill_missing = "zero", time_zero_shift = FALSE,
               x_breaks = c(8, 16, 24, 40), interactive = "ggiraph")
```

</details>

The study source is [Chen, Johnson, Hérissant et al. (2023)](https://doi.org/10.7554/eLife.92899),
BioProject **PRJNA912754**. Published analysis files are pinned to author repository
revision `a375a116bf69160f634a3d6a1b0ecc53ad62d142`. The separate validation receipt
records hashes of the exact report inputs, R source files and final HTML.

## What this supports, and what remains open

**The completed analysis supports fast retrospective reconstruction with very
high agreement on the published pair set.** It does not establish that barbac is
universally most accurate, or that every unmatched pair is an error. The author
pipeline includes additional lane, abundance, chimera and composition filters
that are not reproduced in this comparison.

The next useful investigation is to trace the two abundant unpublished pairs
through the author's filtering and lineage-selection steps. That could explain
a large part of the reference mismatch. Independent labeled controls would be
needed to turn experimental agreement into a direct accuracy measurement.

Pooling all four generations makes this analysis retrospective. No fitness
coefficients, causal effects or performance on future timepoints are inferred.
Open questions are whether unmatched abundant pairs are biological lineages or
artifacts, which publication filters removed them, and how stable the correction
is in an independent assay with different abundance and error patterns.
