---
title: "barbac Studio · Analysis report"
format:
  html:
    embed-resources: true
    theme: cosmo
    toc: true
    code-fold: true
    html-math-method: plain
execute:
  echo: false
  warning: false
  message: false
---

```{r}
x <- readRDS('analysis.rds')
plots <- readRDS('plots.rds')
```

```{=html}
<style>
body{color:#173e3b;background:#fbfaf6;font-family:Arial,sans-serif}
h1,h2,h3{color:#123f3b}h1{font-weight:700;letter-spacing:-1.4px}
.quarto-title-block{border-bottom:3px solid #d5b574;padding-bottom:22px}
table{font-size:.9rem} a{color:#247b70}
</style>
```

## Analysis overview

This report contains a completed clustering run from barbac Studio.
`r if(isTRUE(x$provenance$synthetic)) '**The input is a synthetic demonstration dataset.**' else ''`

The analysis contains **`r format(sum(x$input$counts),big.mark=',')` barcode reads**,
**`r length(unique(x$input$sample))` samples**, and
**`r nrow(x$centroids)` inferred clusters** across
**`r length(unique(x$input$population))` populations**.

Clustering took `r sprintf('%.3f',x$seconds)` seconds on the machine running the app;
this excludes upload, extraction, plotting and report rendering.

## Clustering settings

```{r}
knitr::kable(data.frame(Setting=names(x$settings),Value=unlist(x$settings)),row.names=FALSE)
```

Counts were pooled across all supplied timepoints within each population.
Independent populations were clustered separately. This retrospective analysis
uses information from later timepoints to define the shared cluster memberships.

The experimental Poisson option, if enabled, modifies the LV merge guard and may
merge genuine length variants. It is not an estimated sequencing error rate.

## Cluster statistics

```{r}
knitr::kable(x$stats,digits=6,row.names=FALSE)
```

Singletons have a total count of one. Largest-cluster and top-K fractions use all
barcode counts in the population as their denominator. Cluster counts are inferred
lineages, not known-truth accuracy estimates.

## Lineage trajectories

Frequencies divide each lineage's counts by all supplied barcode counts in that
sample. No abundance cutoff is applied. Missing lineage counts are filled with zero.
Plots require at least two numeric timepoints in a population.

```{r, results='asis'}
if(!length(plots))cat('No trajectory plots are available for this input.\n')
for(p in plots){
  cat('\n### ',htmltools::htmlEscape(p$population),'\n\n',sep='')
  cat('![](',p$file,')\n\n',sep='')
}
```

## Most abundant clusters

The table shows the 30 most abundant clusters. Complete centroids, memberships
and per-sample counts are available in the app's CSV and ZIP downloads.

```{r}
knitr::kable(head(x$centroids[order(-x$centroids$sum_counts),],30),row.names=FALSE)
```

## Reproducibility

Package version: `r x$provenance$package_version`  
Clustering build: `r x$provenance$build_id`  
Created: `r x$provenance$created_utc`  
Normalized input SHA-256: `r x$provenance$input_sha256`

```{r}
cat(jsonlite::toJSON(list(settings=x$settings,provenance=x$provenance),pretty=TRUE,auto_unbox=TRUE,na='null'))
```
